---
title: "Andes Virus Outbreak"
subtitle: "MV Hondius Cruise Ship · 2026"
date: today
date-format: "D MMMM YYYY"
format:
  pdf:
    documentclass: article
    geometry: "margin=2.5cm"
    fontsize: 11pt
    toc: true
    fig-width: 7
    fig-height: 4.5
    keep-tex: false
  html:
    theme: cosmo
    toc: true
    embed-resources: true
    fig-width: 9
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

```{bash}
#| include: false
# git -C Hondius_hantavirus_h2026 checkout main
# git -C Hondius_hantavirus_h2026 pull origin main
```

```{r setup}
#| include: false
library(tidyverse)
library(outbreaker2)
library(ape)
library(epitrix)
library(distcrete)
library(epicontacts)
library(visNetwork)
library(tidygraph)
library(ggraph)
library(gt)
library(o2ools)
library(httr2)
library(scales)
library(ggridges)
library(phangorn)
library(pheatmap)
library(ggtree)
library(coda)
library(furrr)
library(future)

group_pal <- c(passenger = "#1f77b4", crew = "#d62728")
```

```{r linelist}
#| include: false
linelist <- read_csv(
  "Hondius_hantavirus_h2026/data/linelist/2026_hantavirus.csv",
  col_types = cols(
    Gh_ID = col_character(),
    symptom_onset = col_date(),
    outcome_date = col_date(),
    ship_boarded = col_date(),
    left_ship = col_date(),
    confirmation_date = col_date()
  )
) |>
  filter(status %in% c("confirmed", "probable")) |>
  arrange(symptom_onset) |>
  mutate(
    group = case_when(
      passenger..y.n. == "y" ~ "passenger",
      cruise.crew..y.n. == "y" ~ "crew"
    ) |>
      factor(levels = c("passenger", "crew")),
    dates = as.integer(symptom_onset - min(symptom_onset, na.rm = TRUE))
  )
```

```{r sequences}
#| include: false
acc_ids <- linelist |>
  filter(!is.na(accession_id)) |>
  pull(accession_id) |>
  str_trim()

fasta <- "Hondius_hantavirus_h2026/data/sequences/matched_andv_s_alignment.fasta"

request(
  "https://lapis.pathoplexus.org/andv/sample/alignedNucleotideSequences/L"
) |>
  req_body_json(list(accessionVersion = as.list(acc_ids))) |>
  req_method("POST") |>
  req_perform() |>
  resp_body_raw() |>
  writeBin(fasta)

dna <- read.FASTA(fasta)
names(dna) <- linelist$Gh_ID[match(names(dna), str_trim(linelist$accession_id))]
dna <- dna[!is.na(names(dna))]
```

```{r analysis}
#| cache: true
#| include: false
params <- gamma_mucv2shapescale(mu = 18, cv = 6 / 18)
si <- distcrete("gamma", 1, shape = params$shape, scale = params$scale)$d(1:50)

o2_data <- outbreaker_data(
  ids = linelist$Gh_ID,
  dates = linelist$dates,
  dna = dna,
  w_dens = si
)

config <- create_config(
  n_iter = 1e5,
  sample_every = 50,

  max_kappa = 3,
  move_kappa = TRUE,

  init_pi = 0.95,
  move_pi = TRUE,
  prior_pi = c(10, 1),  # Beta(10, 1), concentrated near 1

  find_import = TRUE,
  n_iter_import = 1e4,
  outlier_threshold = 2.5, #default is 5, lower = more aggressive outlier detection

  move_mu = FALSE,
  init_mu  = 1.7e-5    # ANDV: ~3.5e-4 subs/site/year × 18/365 days/generation
)

# Run 4 independent chains in parallel, each with a distinct seed
plan(multisession, workers = 4)

o2_chains <- future_map(
  c(123L, 456L, 789L, 1011L),
  \(s) {
    set.seed(s)
    outbreaker(data = o2_data, config = config) |> filter(step > 10000)
  },
  .options = furrr_options(seed = NULL)
)

plan(sequential)

# Pool all chains; .id = "chain" records which chain each row came from
# Restore outbreaker_chains class so o2ools S3 methods dispatch correctly
o2 <- bind_rows(o2_chains, .id = "chain")
class(o2) <- c("outbreaker_chains", class(o2))
o2_id <- o2ools::identify(o2, ids = linelist$Gh_ID)
entropy_vals <- get_entropy(o2_id)
consensus_tree <- get_consensus(o2_id)
```

::: {.callout-note appearance="minimal"}
**Outbreak summary** · `r format(Sys.Date(), "%d %B %Y")`

`r nrow(linelist)` cases in analysis (`r sum(linelist$status == "confirmed")` confirmed · `r sum(linelist$status == "probable")` probable) · `r sum(!is.na(linelist$outcome) & linelist$outcome == "died")` deaths · `r format(min(linelist$symptom_onset, na.rm=TRUE), "%d %b")` – `r format(max(linelist$symptom_onset, na.rm=TRUE), "%d %b %Y")`
:::

## Background

Andes virus (ANDV) is the only hantavirus documented to transmit person-to-person, in addition to the usual rodent-to-human route. The 2026 outbreak is linked to the **MV Hondius** expedition cruise ship. Cases first appeared in early April 2026; by the time infection was confirmed, passengers and crew had dispersed across Europe, Asia, and the Americas. Whole-genome sequencing confirms ANDV as the causative agent.

## Methods

Transmission trees were inferred with **outbreaker2**, integrating symptom onset dates and genome sequences. Serial interval was modelled as Gamma(mean = 18 d, CV = 1/3). MCMC: 100,000 iterations, thinned every 50 steps, 10% burn-in. L-segment sequences were retrieved from [Pathoplexus](https://pathoplexus.org) for the `r length(dna)` of `r nrow(linelist)` cases with deposited accessions. The S and M segments are identical across all cases; the L segment carries the only two SNPs observed in the outbreak and is therefore the sole source of genetic signal. Remaining cases contribute temporal data only. Because near-zero genetic variation renders the per-site mutation rate ($\mu$) non-identifiable from the data, $\mu$ was fixed at $1.7 \times 10^{-5}$ substitutions/site/generation, derived from the published ANDV substitution rate of ${\sim}3.5 \times 10^{-4}$ subs/site/year and an 18-day generation interval.


## Results

### Cases

```{r case-table}
#| tbl-cap: "Confirmed and probable cases."
tbl_data <- linelist |>
  transmute(
    `Case ID`    = Gh_ID,
    Status       = str_to_title(status),
    Role         = str_to_title(as.character(group)),
    `Age (yr)`   = as.character(age),
    Sex          = str_to_title(sex),
    Nationality  = str_to_title(nationality),
    `Onset date` = if_else(!is.na(symptom_onset),
                           format(symptom_onset, "%d %b %Y"), "—"),
    Outcome      = case_when(
                     outcome == "died" ~ "Fatal",
                     !is.na(outcome)   ~ str_to_title(outcome),
                     TRUE              ~ "—"),
    Sequence     = if_else(!is.na(accession_id), str_trim(accession_id), "—")
  )

if (knitr::is_latex_output()) {
  knitr::kable(tbl_data, booktabs = TRUE, linesep = "")
} else {
  tbl_data |>
    gt() |>
    tab_style(
      style     = cell_text(weight = "bold", color = "#c0392b"),
      locations = cells_body(rows = Outcome == "Fatal", columns = Outcome)
    ) |>
    tab_style(
      style     = list(cell_fill(color = "#2c3e50"),
                       cell_text(color = "white", weight = "bold")),
      locations = cells_column_labels()
    ) |>
    tab_options(table.width = pct(100), table.font.size = 13,
                row.striping.include_table_body = TRUE)
}
```

### Genomic Analysis

```{r snp-matrix}
#| fig-cap: "Pairwise SNP distance matrix for sequenced cases (L segment). Values are raw nucleotide differences. Near-zero distances confirm a near-clonal cluster consistent with a single spillover event."
#| fig-height: 3.5

snp_dist <- ape::dist.dna(dna, model = "N", pairwise.deletion = TRUE) |>
  as.matrix()

pheatmap(
  snp_dist,
  color            = colorRampPalette(c("#f7fbff", "#21b528"))(50),
  display_numbers  = TRUE,
  number_format    = "%.0f",
  number_color     = "black",
  cluster_rows     = TRUE,
  cluster_cols     = TRUE,
  treeheight_row   = 20,
  treeheight_col   = 20,
  border_color     = "white",
  fontsize         = 11,
  legend           = FALSE,
  main             = ""
)
```

### Model Diagnostics

```{r diagnostics-trace}
#| fig-cap: "MCMC trace plots for scalar parameters (post-burnin). Stationary, well-mixed traces indicate convergence."
#| fig-height: 7
param_labels <- c(
  post = "Log posterior",
  mu = "Mutation rate (μ)",
  pi = "Reporting rate (π)"
)

o2 |>
  mutate(chain = paste("Chain", chain)) |>
  select(chain, step, all_of(names(param_labels))) |>
  pivot_longer(-c(chain, step), names_to = "param", values_to = "value") |>
  mutate(param = factor(param_labels[param], levels = param_labels)) |>
  ggplot(aes(x = step, y = value, colour = chain)) +
  geom_line(linewidth = 0.25, alpha = 0.8) +
  facet_wrap(~param, scales = "free_y", ncol = 1) +
  scale_colour_brewer(palette = "Set1", name = NULL) +
  labs(x = "MCMC iteration", y = NULL) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "#f0f0f0"),
    legend.position = "bottom"
  )
```

```{r diagnostics-gelman}
#| tbl-cap: "Convergence diagnostics across 4 independent chains. R-hat < 1.01 and ESS > 400 indicate good convergence and mixing."
# R-hat and ESS only for actual sampled parameters.
# mu is fixed (move_mu = FALSE in config), so R-hat/ESS are undefined (0/0 → NaN)
# and it is intentionally omitted from the table.
diag_params <- c(pi = "Reporting rate (π)")

mcmc_chains <- split(o2, o2$chain) |>
  lapply(\(df) coda::mcmc(as.matrix(df[names(diag_params)]))) |>
  coda::mcmc.list()

gd  <- coda::gelman.diag(mcmc_chains, confidence = 0.95,
                          transform = FALSE, autoburnin = FALSE,
                          multivariate = FALSE)
ess <- coda::effectiveSize(mcmc_chains)

as.data.frame(gd$psrf) |>
  setNames(c("R-hat", "97.5% CI")) |>
  rownames_to_column("param") |>
  mutate(Parameter = diag_params[param],
         ESS       = round(ess[param]),
         .keep = "unused") |>
  relocate(Parameter) |>
  gt() |>
  fmt_number(columns = c("R-hat", "97.5% CI"), decimals = 3) |>
  fmt_integer(columns = ESS) |>
  tab_style(
    style = cell_fill(color = "#d4edda"),
    locations = cells_body(rows = `R-hat` < 1.01)
  ) |>
  tab_style(
    style = cell_fill(color = "#fff3cd"),
    locations = cells_body(rows = `R-hat` >= 1.01 & `R-hat` < 1.05)
  ) |>
  tab_style(
    style = cell_fill(color = "#f8d7da"),
    locations = cells_body(rows = `R-hat` >= 1.05)
  ) |>
  tab_style(
    style = list(
      cell_fill(color = "#2c3e50"),
      cell_text(color = "white", weight = "bold")
    ),
    locations = cells_column_labels()
  ) |>
  tab_footnote(paste0(
    "R-hat: Green < 1.01 · Yellow 1.01–1.05 · Red ≥ 1.05. ",
    "ESS: effective sample size out of ", nrow(o2), " total posterior samples ",
    "(", n_distinct(o2$chain), " chains × ",
    nrow(o2) / n_distinct(o2$chain), " samples per chain)."
  )) |>
  tab_options(table.width = pct(65), table.font.size = 13)
```

### Transmission Tree

```{r fig-network-static}
#| fig-cap: "Consensus transmission tree (coloured edges), where each case is assigned its most frequently inferred ancestor across the posterior sample. Grey dashed arrows show the posterior support for direct transmission from case 1 to all cases not already linked to it in the consensus tree, consistent with a superspreading event."
#| fig-height: 6
draw_key_arrow <- function(data, params, size) {
  grid::segmentsGrob(
    x0 = 0.1,
    x1 = 0.85,
    y0 = 0.5,
    y1 = 0.5,
    arrow = grid::arrow(length = unit(2, "mm"), type = "closed"),
    gp = grid::gpar(
      col = "grey25",
      fill = "grey25",
      lwd = (data$edge_width %||% 0.5) * ggplot2::.pt
    )
  )
}

epi <- make_epicontacts(
  linelist = linelist,
  contacts = subset(consensus_tree, !is.na(from)),
  directed = TRUE
)

g <- epicontacts:::as.igraph.epicontacts(epi) |> as_tbl_graph()

# Reingold-Tilford: designed for rooted trees, guarantees no overlap
roots <- which(igraph::degree(g, mode = "in") == 0)
tree_layout <- igraph::layout_as_tree(g, root = roots)
rownames(tree_layout) <- igraph::V(g)$name

layout_data <- create_layout(g, layout = "kk")
layout_data$x <- as.numeric(layout_data$symptom_onset)
layout_data$y <- tree_layout[layout_data$name, 1] # tree breadth → y axis

# Supporting edges: case "1" → every node not already linked in the consensus tree
already_linked <- subset(consensus_tree, from == "1" & !is.na(from))$to
ss_targets <- setdiff(linelist$Gh_ID, c("1", already_linked))
src <- layout_data[layout_data$name == "1", c("x", "y")]
ss_segs <- data.frame(
  x    = src$x, y    = src$y,
  xend = layout_data$x[match(ss_targets, layout_data$name)],
  yend = layout_data$y[match(ss_targets, layout_data$name)],
  freq = sapply(ss_targets, \(t) mean(o2_id[[paste0("alpha_", t)]] == "1", na.rm = TRUE))
) |> mutate(xmid = (x + xend) / 2, ymid = (y + yend) / 2)

ggraph(layout_data) +
  geom_segment(
    data = ss_segs,
    aes(x = x, y = y, xend = xend, yend = yend),
    colour = "grey60", linetype = "dashed",
    arrow = arrow(length = unit(2.5, "mm"), type = "closed"),
    inherit.aes = FALSE
  ) +
  geom_text(
    data = ss_segs,
    aes(x = xmid, y = ymid, label = sprintf("%.2f", freq)),
    colour = "grey50", size = 3, inherit.aes = FALSE
  ) +
  geom_edge_link(
    aes(
      edge_width = frequency,
      color = .N()$group[from],
      label = sprintf("%.2f", frequency)
    ),
    arrow = arrow(length = unit(2.5, "mm")),
    end_cap = circle(3, "mm"),
    angle_calc = "along",
    label_dodge = unit(2.5, "mm"),
    label_size = 3,
    label_colour = "black",
    key_glyph = draw_key_arrow
  ) +
  geom_node_point(aes(fill = group), shape = 21, colour = "black", size = 5) +
  geom_node_text(
    aes(label = name),
    repel = TRUE,
    size = 3,
    nudge_y = 0.1,
    bg.color = "white",
    bg.r = 0.1
  ) +
  scale_edge_width(
    "Posterior support",
    range = c(0.1, 1),
    breaks = c(0.8, 0.9, 1)
  ) +
  scale_fill_manual(
    NULL,
    values = group_pal,
    breaks = names(group_pal),
    labels = c(passenger = "Passenger", crew = "Crew")
  ) +
  scale_edge_colour_manual(values = group_pal, guide = "none") +
  guides(
    fill = guide_legend(override.aes = list(size = 4), order = 1),
    edge_width = guide_legend(order = 2)
  ) +
  scale_x_continuous(breaks = breaks_width(3), labels = \(x) {
    format(as.Date(x, origin = "1970-01-01"), "%b %d")
  }) +
  labs(x = "Symptom onset date", y = "") +
  theme_bw(base_size = 13) +
  theme(
    axis.line.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "bottom"
  )
```

### Ancestry Uncertainty

```{r entropy-plot}
#| fig-cap: "Shannon entropy of each case's inferred source. Entropy = 0 means a single unambiguous infector."
#| fig-height: 4
onset_order <- linelist |> arrange(desc(symptom_onset)) |> pull(Gh_ID)

tibble(
  case = gsub("alpha_", "", names(entropy_vals)),
  entropy = entropy_vals
) |>
  mutate(case = factor(case, levels = onset_order)) |>
  ggplot(aes(x = entropy, y = case, fill = entropy)) +
  geom_col() +
  scale_fill_gradient(low = "#caced1", high = "#383939", guide = "none") +
  labs(x = "Shannon entropy", y = "Case ID") +
  theme_bw(base_size = 13) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )
```

### Dates of Infection

```{r infection-dates}
#| fig-cap: "Posterior distribution of inferred infection dates per case. Width reflects uncertainty; cases ordered by mean inferred infection date."
#| fig-height: 6
ref_date <- min(linelist$symptom_onset, na.rm = TRUE)

o2 |>
  select(starts_with("t_inf")) |>
  pivot_longer(everything(), names_to = "col", values_to = "t_inf") |>
  mutate(
    idx = as.integer(str_extract(col, "\\d+")),
    case = fct_reorder(linelist$Gh_ID[idx], t_inf, mean, .desc = TRUE),
    grp = linelist$group[idx],
    date = ref_date + t_inf
  ) |>
  ggplot(aes(x = date, y = case, fill = grp)) +
  geom_density_ridges(alpha = 0.75, colour = "white", linewidth = 0.4) +
  scale_fill_manual(
    values = group_pal,
    na.value = "grey70",
    labels = c(passenger = "Passenger", crew = "Crew"),
    name = NULL
  ) +
  scale_x_date(
    date_breaks = "7 days",
    date_labels = "%d %b",
    expand = expansion(add = 3)
  ) +
  labs(x = "Inferred date of infection", y = NULL) +
  theme_bw(base_size = 13) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 30, hjust = 1)
  )
```

### Imports and Unobserved Transmission

```{r import-kappa}
#| fig-cap: "Posterior probability of each case being imported, directly linked, or connected via unobserved intermediate cases (κ)."
#| fig-height: 5
n_cases <- nrow(linelist)

# Stack alpha and kappa column-wise: n_samples × n_cases → long format
plot_df <- tibble(
  case = rep(linelist$Gh_ID, each = nrow(o2)),
  alpha = as.vector(as.matrix(o2[paste0("alpha_", seq_len(n_cases))])),
  kappa = as.vector(as.matrix(o2[paste0("kappa_", seq_len(n_cases))]))
) |>
  mutate(
    state = case_when(
      is.na(alpha) ~ "Imported",
      kappa == 1 ~ "Direct (κ = 1)",
      kappa == 2 ~ "1 unobserved (κ = 2)",
      TRUE ~ "2+ unobserved (κ ≥ 3)"
    ) |>
      factor(
        levels = c(
          "Imported",
          "Direct (κ = 1)",
          "1 unobserved (κ = 2)",
          "2+ unobserved (κ ≥ 3)"
        )
      )
  ) |>
  count(case, state) |>
  group_by(case) |>
  mutate(prob = n / sum(n)) |>
  ungroup() |>
  complete(case, state, fill = list(n = 0, prob = 0))

case_order <- linelist |>
  arrange(desc(symptom_onset)) |>
  pull(Gh_ID)

plot_df |>
  mutate(case = factor(case, levels = case_order)) |>
  ggplot(aes(x = prob, y = case, fill = state)) +
  geom_col() +
  scale_x_continuous(
    labels = percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.02))
  ) +
  scale_fill_manual(
    values = c(
      "Imported" = "#e6550d",
      "Direct (κ = 1)" = "#525252",
      "1 unobserved (κ = 2)" = "#817f7f",
      "2+ unobserved (κ ≥ 3)" = "#c4c3c3"
    ),
    name = NULL
  ) +
  labs(x = "Posterior frequency", y = NULL) +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

### Reproduction Number

```{r offspring}
#| fig-cap: "Posterior distribution of the offspring distribution."
#| fig-height: 4
get_Ri(o2_id) |>
  mutate(step = row_number()) |>
  pivot_longer(-step, names_to = "name", values_to = "Ri") |>
  group_by(step, Ri) |>
  summarise(n = n(), .groups = "drop_last") |>
  mutate(percent = n / sum(n) * 100) |>
  ungroup() |>
  group_by(Ri) |>
  summarise(
    mean = mean(percent),
    lower = quantile(percent, 0.025),
    upper = quantile(percent, 0.975),
    .groups = "drop"
  ) |>
  ggplot(aes(x = Ri, y = mean)) +
  geom_segment(
    aes(x = Ri, xend = Ri, y = 0, yend = mean),
    colour = "grey60",
    size = 0.5
  ) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
  theme_bw()
```

## Data

Data available under [CC BY 4.0](https://creativecommons.org/licenses/by/4.0/) at [kraemer-lab/Hondius_hantavirus_h2026](https://github.com/kraemer-lab/Hondius_hantavirus_h2026). Sequences retrieved from [Pathoplexus](https://pathoplexus.org). Report source: `report.qmd`.
