---
title: "Andes Virus Outbreak"
subtitle: "MV Hondius Cruise Ship · 2026"
date: today
date-format: "D MMMM YYYY"
format:
  pdf:
    documentclass: article
    geometry: "margin=2.5cm"
    fontsize: 11pt
    toc: true
    fig-width: 7
    fig-height: 4.5
    keep-tex: false
  html:
    theme: cosmo
    toc: true
    embed-resources: true
    fig-width: 9
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

```{bash}
#| include: false
git -C Hondius_hantavirus_h2026 pull
```

```{r setup}
#| include: false
library(tidyverse)
library(outbreaker2)
library(ape)
library(epitrix)
library(distcrete)
library(epicontacts)
library(visNetwork)
library(tidygraph)
library(ggraph)
library(gt)
library(o2ools)
library(httr2)
library(scales)
library(ggridges)
library(coda)
library(furrr)
library(future)

group_pal <- c(passenger = "#1f77b4", crew = "#d62728")
```

```{r linelist}
#| include: false
linelist <- read_csv(
  "Hondius_hantavirus_h2026/data/linelist/2026_hantavirus.csv",
  col_types = cols(
    Gh_ID = col_character(),
    symptom_onset = col_date(),
    outcome_date = col_date(),
    ship_boarded = col_date(),
    left_ship = col_date(),
    confirmation_date = col_date()
  )
) |>
  filter(status %in% c("confirmed", "probable")) |>
  arrange(symptom_onset) |>
  mutate(
    group = case_when(
      passenger..y.n. == "y" ~ "passenger",
      cruise.crew..y.n. == "y" ~ "crew"
    ) |>
      factor(levels = c("passenger", "crew")),
    dates = as.integer(symptom_onset - min(symptom_onset, na.rm = TRUE))
  )
```

```{r sequences}
#| include: false
acc_ids <- linelist |>
  filter(!is.na(accession_id)) |>
  pull(accession_id) |>
  str_trim()

fasta <- "Hondius_hantavirus_h2026/data/sequences/matched_andv_s_alignment.fasta"

request(
  "https://lapis.pathoplexus.org/andv/sample/alignedNucleotideSequences/L"
) |>
  req_body_json(list(accessionVersion = as.list(acc_ids))) |>
  req_method("POST") |>
  req_perform() |>
  resp_body_raw() |>
  writeBin(fasta)

dna <- read.FASTA(fasta)
names(dna) <- linelist$Gh_ID[match(names(dna), str_trim(linelist$accession_id))]
dna <- dna[!is.na(names(dna))]
```

```{r analysis}
#| cache: true
#| include: false
params <- gamma_mucv2shapescale(mu = 18, cv = 6 / 18)
si <- distcrete("gamma", 1, shape = params$shape, scale = params$scale)$d(1:50)

o2_data <- outbreaker_data(
  ids = linelist$Gh_ID,
  dates = linelist$dates,
  dna = dna,
  w_dens = si
)

config <- create_config(
  n_iter = 1e5,
  sample_every = 50,
  max_kappa = 5,
  move_kappa = TRUE,
  init_pi = 0.9,
  move_pi = TRUE,
  find_import = TRUE,
  n_iter_import = 1e4,
  outlier_threshold = 2.5 #default is 5, lower = more aggressive outlier detection
)

# Run 4 independent chains in parallel, each with a distinct seed
plan(multisession, workers = 4)

o2_chains <- future_map(
  c(123L, 456L, 789L, 1011L),
  \(s) {
    set.seed(s)
    outbreaker(data = o2_data, config = config) |> filter(step > 10000)
  },
  .options = furrr_options(seed = NULL)
)

plan(sequential)

# Pool all chains; .id = "chain" records which chain each row came from
o2 <- bind_rows(o2_chains, .id = "chain")
o2_id <- identify(o2, ids = linelist$Gh_ID)
entropy_vals <- get_entropy(o2_id)
consensus_tree <- get_consensus(o2_id)
```

::: {.callout-note appearance="minimal"}
**Outbreak summary** · `r format(Sys.Date(), "%d %B %Y")`

`r nrow(linelist)` cases in analysis (`r sum(linelist$status == "confirmed")` confirmed · `r sum(linelist$status == "probable")` probable) · `r sum(!is.na(linelist$outcome) & linelist$outcome == "died")` deaths · `r format(min(linelist$symptom_onset, na.rm=TRUE), "%d %b")` – `r format(max(linelist$symptom_onset, na.rm=TRUE), "%d %b %Y")`
:::

## Background

Andes virus (ANDV) is the only hantavirus documented to transmit person-to-person, in addition to the usual rodent-to-human route. The 2026 outbreak is linked to the **MV Hondius** expedition cruise ship. Cases first appeared in early April 2026; by the time infection was confirmed, passengers and crew had dispersed across Europe, Asia, and the Americas. S-segment whole-genome sequencing confirms ANDV as the causative agent.

## Methods

Transmission trees were inferred with **outbreaker2**, integrating symptom onset dates and pairwise S-segment genetic distances. Serial interval was modelled as Gamma(mean = 18 d, CV = 1/3). MCMC: 100,000 iterations, thinned every 50 steps, 10% burn-in. L-segment sequences were retrieved from [Pathoplexus](https://pathoplexus.org) for the `r length(dna)` of `r nrow(linelist)` cases with deposited accessions. The S and M segments are identical across all cases; the L segment carries the only two SNPs observed in the outbreak and is therefore the sole source of genetic signal. Remaining cases contribute temporal data only.

## Results

### Cases

```{r case-table}
#| tbl-cap: "Confirmed and probable cases."
linelist |>
  transmute(
    `Case ID` = Gh_ID,
    Status = str_to_title(status),
    Role = str_to_title(as.character(group)),
    `Age (yr)` = as.character(age),
    Sex = str_to_title(sex),
    Nationality = str_to_title(nationality),
    `Onset date` = if_else(
      !is.na(symptom_onset),
      format(symptom_onset, "%d %b %Y"),
      "—"
    ),
    Outcome = case_when(
      outcome == "died" ~ "Fatal",
      !is.na(outcome) ~ str_to_title(outcome),
      TRUE ~ "—"
    ),
    Sequence = if_else(!is.na(accession_id), str_trim(accession_id), "—")
  ) |>
  gt() |>
  tab_style(
    style = cell_text(weight = "bold", color = "#c0392b"),
    locations = cells_body(rows = Outcome == "Fatal", columns = Outcome)
  ) |>
  tab_style(
    style = list(
      cell_fill(color = "#2c3e50"),
      cell_text(color = "white", weight = "bold")
    ),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.width = pct(100),
    table.font.size = 13,
    row.striping.include_table_body = TRUE
  )
```

### Model Diagnostics

```{r diagnostics-trace}
#| fig-cap: "MCMC trace plots for scalar parameters (post-burnin). Stationary, well-mixed traces indicate convergence."
#| fig-height: 7
param_labels <- c(
  post = "Log posterior",
  like = "Log likelihood",
  mu = "Mutation rate (μ)",
  pi = "Reporting rate (π)"
)

o2 |>
  mutate(chain = paste("Chain", chain)) |>
  select(chain, step, all_of(names(param_labels))) |>
  pivot_longer(-c(chain, step), names_to = "param", values_to = "value") |>
  mutate(param = factor(param_labels[param], levels = param_labels)) |>
  ggplot(aes(x = step, y = value, colour = chain)) +
  geom_line(linewidth = 0.25, alpha = 0.8) +
  facet_wrap(~param, scales = "free_y", ncol = 1) +
  scale_colour_brewer(palette = "Set1", name = NULL) +
  labs(x = "MCMC iteration", y = NULL) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "#f0f0f0"),
    legend.position = "bottom"
  )
```

```{r diagnostics-gelman}
#| tbl-cap: "Gelman-Rubin convergence diagnostic across 4 independent chains. R-hat < 1.01 indicates good convergence; values ≥ 1.05 suggest insufficient mixing."
# Split pooled o2 back by chain for Gelman-Rubin
mcmc_chains <- split(o2, o2$chain) |>
  lapply(\(df) coda::mcmc(as.matrix(df[names(param_labels)]))) |>
  coda::mcmc.list()

gd <- coda::gelman.diag(
  mcmc_chains,
  confidence = 0.95,
  transform = FALSE,
  autoburnin = FALSE,
  multivariate = FALSE
)

as.data.frame(gd$psrf) |>
  setNames(c("R-hat", "97.5% CI")) |>
  rownames_to_column("param") |>
  mutate(Parameter = param_labels[param], .keep = "unused") |>
  relocate(Parameter) |>
  gt() |>
  fmt_number(columns = c("R-hat", "97.5% CI"), decimals = 3) |>
  tab_style(
    style = cell_fill(color = "#d4edda"),
    locations = cells_body(rows = `R-hat` < 1.01)
  ) |>
  tab_style(
    style = cell_fill(color = "#fff3cd"),
    locations = cells_body(rows = `R-hat` >= 1.01 & `R-hat` < 1.05)
  ) |>
  tab_style(
    style = cell_fill(color = "#f8d7da"),
    locations = cells_body(rows = `R-hat` >= 1.05)
  ) |>
  tab_style(
    style = list(
      cell_fill(color = "#2c3e50"),
      cell_text(color = "white", weight = "bold")
    ),
    locations = cells_column_labels()
  ) |>
  tab_footnote("Green < 1.01 · Yellow 1.01–1.05 · Red ≥ 1.05") |>
  tab_options(table.width = pct(55), table.font.size = 13)
```

### Transmission Tree

```{r fig-network-static}
#| fig-cap: "Consensus transmission tree, where each case is assigned its most frequently inferred ancestor across the posterior sample."
#| fig-height: 6
draw_key_arrow <- function(data, params, size) {
  grid::segmentsGrob(
    x0 = 0.1,
    x1 = 0.85,
    y0 = 0.5,
    y1 = 0.5,
    arrow = grid::arrow(length = unit(2, "mm"), type = "closed"),
    gp = grid::gpar(
      col = "grey25",
      fill = "grey25",
      lwd = (data$edge_width %||% 0.5) * ggplot2::.pt
    )
  )
}

epi <- make_epicontacts(
  linelist = linelist,
  contacts = subset(consensus_tree, !is.na(from)),
  directed = TRUE
)

g <- epicontacts:::as.igraph.epicontacts(epi) |> as_tbl_graph()

# Reingold-Tilford: designed for rooted trees, guarantees no overlap
roots <- which(igraph::degree(g, mode = "in") == 0)
tree_layout <- igraph::layout_as_tree(g, root = roots)
rownames(tree_layout) <- igraph::V(g)$name

layout_data <- create_layout(g, layout = "kk")
layout_data$x <- as.numeric(layout_data$symptom_onset)
layout_data$y <- tree_layout[layout_data$name, 1] # tree breadth → y axis

ggraph(layout_data) +
  geom_edge_link(
    aes(
      edge_width = frequency,
      color = .N()$group[from],
      label = sprintf("%.2f", frequency)
    ),
    arrow = arrow(length = unit(2.5, "mm")),
    end_cap = circle(3, "mm"),
    angle_calc = "along",
    label_dodge = unit(2.5, "mm"),
    label_size = 3,
    label_colour = "black",
    key_glyph = draw_key_arrow
  ) +
  geom_node_point(aes(fill = group), shape = 21, colour = "black", size = 5) +
  geom_node_text(
    aes(label = name),
    repel = TRUE,
    size = 3,
    nudge_y = 0.1,
    bg.color = "white",
    bg.r = 0.1
  ) +
  scale_edge_width(
    "Posterior support",
    range = c(0.1, 1),
    breaks = c(0.8, 0.9, 1)
  ) +
  scale_fill_manual(
    NULL,
    values = group_pal,
    breaks = names(group_pal),
    labels = c(passenger = "Passenger", crew = "Crew")
  ) +
  scale_edge_colour_manual(values = group_pal, guide = "none") +
  guides(
    fill = guide_legend(override.aes = list(size = 4), order = 1),
    edge_width = guide_legend(order = 2)
  ) +
  scale_x_continuous(breaks = breaks_width(3), labels = \(x) {
    format(as.Date(x, origin = "1970-01-01"), "%b %d")
  }) +
  labs(x = "Symptom onset date", y = "") +
  theme_bw(base_size = 13) +
  theme(
    axis.line.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "bottom"
  )
```

### Dates of Infection

```{r infection-dates}
#| fig-cap: "Posterior distribution of inferred infection dates per case. Width reflects uncertainty; cases ordered by mean inferred infection date."
#| fig-height: 6
ref_date <- min(linelist$symptom_onset, na.rm = TRUE)

o2 |>
  select(starts_with("t_inf")) |>
  pivot_longer(everything(), names_to = "col", values_to = "t_inf") |>
  mutate(
    idx = as.integer(str_extract(col, "\\d+")),
    case = fct_reorder(linelist$Gh_ID[idx], t_inf, mean, .desc = TRUE),
    grp = linelist$group[idx],
    date = ref_date + t_inf
  ) |>
  ggplot(aes(x = date, y = case, fill = grp)) +
  geom_density_ridges(alpha = 0.75, colour = "white", linewidth = 0.4) +
  scale_fill_manual(
    values = group_pal,
    na.value = "grey70",
    labels = c(passenger = "Passenger", crew = "Crew"),
    name = NULL
  ) +
  scale_x_date(
    date_breaks = "7 days",
    date_labels = "%d %b",
    expand = expansion(add = 3)
  ) +
  labs(x = "Inferred date of infection", y = NULL) +
  theme_bw(base_size = 13) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 30, hjust = 1)
  )
```

### Imports and Unobserved Transmission

```{r import-kappa}
#| fig-cap: "Posterior probability of each case being imported, directly linked, or connected via unobserved intermediate cases (κ)."
#| fig-height: 5
n_cases <- nrow(linelist)

# Stack alpha and kappa column-wise: n_samples × n_cases → long format
plot_df <- tibble(
  case = rep(linelist$Gh_ID, each = nrow(o2)),
  alpha = as.vector(as.matrix(o2[paste0("alpha_", seq_len(n_cases))])),
  kappa = as.vector(as.matrix(o2[paste0("kappa_", seq_len(n_cases))]))
) |>
  mutate(
    state = case_when(
      is.na(alpha) ~ "Imported",
      kappa == 1 ~ "Direct (κ = 1)",
      kappa == 2 ~ "1 unobserved (κ = 2)",
      TRUE ~ "2+ unobserved (κ ≥ 3)"
    ) |>
      factor(
        levels = c(
          "Imported",
          "Direct (κ = 1)",
          "1 unobserved (κ = 2)",
          "2+ unobserved (κ ≥ 3)"
        )
      )
  ) |>
  count(case, state) |>
  group_by(case) |>
  mutate(prob = n / sum(n)) |>
  ungroup() |>
  complete(case, state, fill = list(n = 0, prob = 0))

case_order <- linelist |>
  arrange(desc(symptom_onset)) |>
  pull(Gh_ID)

plot_df |>
  mutate(case = factor(case, levels = case_order)) |>
  ggplot(aes(x = prob, y = case, fill = state)) +
  geom_col() +
  scale_x_continuous(
    labels = percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.02))
  ) +
  scale_fill_manual(
    values = c(
      "Imported" = "#e6550d",
      "Direct (κ = 1)" = "#525252",
      "1 unobserved (κ = 2)" = "#817f7f",
      "2+ unobserved (κ ≥ 3)" = "#c4c3c3"
    ),
    name = NULL
  ) +
  labs(x = "Posterior frequency", y = NULL) +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

### Ancestry Uncertainty

```{r entropy-plot}
#| fig-cap: "Shannon entropy of each case's inferred source. Entropy = 0 means a single unambiguous infector."
#| fig-height: 4
onset_order <- linelist |> arrange(desc(symptom_onset)) |> pull(Gh_ID)

tibble(
  case = gsub("alpha_", "", names(entropy_vals)),
  entropy = entropy_vals
) |>
  mutate(case = factor(case, levels = onset_order)) |>
  ggplot(aes(x = entropy, y = case, fill = entropy)) +
  geom_col() +
  scale_fill_gradient(low = "#caced1", high = "#383939", guide = "none") +
  labs(x = "Shannon entropy", y = "Case ID") +
  theme_bw(base_size = 13) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )
```

### Reproduction Number

```{r offspring}
#| fig-cap: "Posterior distribution of the offspring distribution."
#| fig-height: 4
get_Ri(o2_id) |>
  mutate(step = row_number()) |>
  pivot_longer(-step, names_to = "name", values_to = "Ri") |>
  group_by(step, Ri) |>
  summarise(n = n(), .groups = "drop_last") |>
  mutate(percent = n / sum(n) * 100) |>
  ungroup() |>
  group_by(Ri) |>
  summarise(
    mean = mean(percent),
    lower = quantile(percent, 0.025),
    upper = quantile(percent, 0.975),
    .groups = "drop"
  ) |>
  ggplot(aes(x = Ri, y = mean)) +
  geom_segment(
    aes(x = Ri, xend = Ri, y = 0, yend = mean),
    colour = "grey60",
    size = 0.5
  ) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
  theme_bw()
```

## Data

Data available under [CC BY 4.0](https://creativecommons.org/licenses/by/4.0/) at [kraemer-lab/Hondius_hantavirus_h2026](https://github.com/kraemer-lab/Hondius_hantavirus_h2026). Sequences retrieved from [Pathoplexus](https://pathoplexus.org). Report source: `report.qmd`.
