---
title: "Andes Virus Outbreak"
subtitle: "MV Hondius Cruise Ship · 2026"
date: today
date-format: "D MMMM YYYY"
format:
  pdf:
    documentclass: article
    geometry: "margin=2.5cm"
    fontsize: 11pt
    toc: true
    fig-width: 7
    fig-height: 4.5
    keep-tex: false
  html:
    theme: cosmo
    toc: true
    embed-resources: true
    fig-width: 9
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

```{bash}
#| include: false
git -C Hondius_hantavirus_h2026 checkout main
git -C Hondius_hantavirus_h2026 pull origin main
```

```{r setup}
#| include: false
library(tidyverse)
library(janitor)
library(outbreaker2)
library(ape)
library(epitrix)
library(distcrete)
library(epicontacts)
library(visNetwork)
library(tidygraph)
library(ggraph)
library(gt)
library(o2ools)
library(httr2)
library(scales)
library(ggridges)
library(phangorn)
library(pheatmap)
library(ggtree)
library(coda)
library(furrr)
library(future)
library(pipetime)

group_pal <- c(passenger = "#1f77b4", crew = "#d62728")
group_labels <- c(passenger = "Passenger", crew = "Crew")
```

```{r linelist}
#| include: false
date_cols <- c(
  "date_onset",
  "date_confirmation",
  "date_hospitalisation",
  "date_death",
  "ship_board_date",
  "ship_disembark_date"
)

linelist <- read_csv(
  "Hondius_hantavirus_h2026/data/linelist/2026_hantavirus.csv",
  col_types = cols(.default = col_character())
) |>
  clean_names() |>
  filter(case_status %in% c("confirmed", "probable")) |>
  mutate(
    across(all_of(date_cols), mdy),
    best_date = pmin(
      date_onset,
      date_confirmation,
      date_hospitalisation,
      na.rm = TRUE
    ),
    dates = as.integer(best_date - min(best_date, na.rm = TRUE)),
    group = factor(
      if_else(cruise_passenger_guest == "Y", "passenger", "crew"),
      levels = c("passenger", "crew")
    )
  ) |>
  select(
    gh_id,
    case_status,
    accession_id,
    group,
    dates,
    best_date,
    ship_board_date,
    ship_disembark_date,
    date_death,
    contact_gh_id,
    contact_setting,
    age,
    gender
  ) |>
  arrange(dates)

ref_date <- min(linelist$best_date, na.rm = TRUE)

ctd_timed <- linelist |>
  filter(!is.na(ship_board_date)) |>
  transmute(
    id = gh_id,
    place = contact_setting,
    start = as.integer(ship_board_date - ref_date),
    end = as.integer(
      pmin(ship_disembark_date, date_death, na.rm = TRUE) - ref_date
    )
  )
```

```{r sequences}
#| include: false
acc_ids <- linelist |>
  filter(!is.na(accession_id)) |>
  pull(accession_id) |>
  str_trim()

fasta <- "Hondius_hantavirus_h2026/data/sequences/matched_andv_s_alignment.fasta"

request(
  "https://lapis.pathoplexus.org/andv/sample/alignedNucleotideSequences/L"
) |>
  req_body_json(list(accessionVersion = as.list(acc_ids))) |>
  req_method("POST") |>
  req_perform() |>
  resp_body_raw() |>
  writeBin(fasta)

dna <- read.FASTA(fasta)
names(dna) <- linelist$gh_id[match(names(dna), str_trim(linelist$accession_id))]
dna <- dna[!is.na(names(dna))]
```

```{r analysis}
#| cache: true
#| include: false
params <- gamma_mucv2shapescale(mu = 18, cv = 6 / 18)
si <- distcrete("gamma", 1, shape = params$shape, scale = params$scale)$d(1:50)

o2_data <- outbreaker_data(
  ids = linelist$gh_id,
  dates = linelist$dates,
  dna = dna,
  ctd_timed = ctd_timed,
  w_dens = si,
  p_trans = list(matrix(1, 1, 1))
)

config <- create_config(
  n_iter = 1e5,
  sample_every = 50,

  # unobserved intermediates
  max_kappa = 3,
  move_kappa = TRUE,

  # reporting rate
  init_pi = 0.95,
  move_pi = TRUE,
  prior_pi = c(10, 1),

  # contact reporting coverage
  init_eps = 1,
  move_eps = FALSE,

  # place persitence
  init_tau = 1,
  move_tau = FALSE,

  # importations
  find_import = TRUE,
  n_iter_import = 1e4,
  outlier_threshold = 2.5, #default is 5, lower = more aggressive outlier detection

  # mutation rate
  move_mu = FALSE,
  init_mu = 1.7e-5 # ANDV: ~3.5e-4 subs/site/year × 18/365 days/generation
)


plan(multisession, workers = 4)
o2_chains <- c(123L, 456L, 789L, 1011L) |>
  future_map(
    \(s) {
      set.seed(s)
      outbreaker(data = o2_data, config = config) |> filter(step > 10000)
    },
    .options = furrr_options(seed = NULL)
  ) |>
  time_pipe("outbreaker2 chains")
plan(sequential)

o2 <- bind_rows(o2_chains, .id = "chain")
class(o2) <- c("outbreaker_chains", class(o2))
o2_id <- o2ools::identify(o2, ids = linelist$gh_id)
entropy_vals <- get_entropy(o2_id)
consensus_tree <- get_consensus(o2_id)
```

::: {.callout-note appearance="minimal"}
**Outbreak summary** · `r format(Sys.Date(), "%d %B %Y")`

`r nrow(linelist)` cases in analysis (`r sum(linelist$case_status == "confirmed")` confirmed · `r sum(linelist$case_status == "probable")` probable) · `r sum(!is.na(linelist$date_death))` deaths · `r format(min(linelist$best_date, na.rm=TRUE), "%d %b")` – `r format(max(linelist$best_date, na.rm=TRUE), "%d %b %Y")`
:::

## Background

Andes virus (ANDV) is the only hantavirus documented to transmit person-to-person, in addition to the usual rodent-to-human route. The 2026 outbreak is linked to the **MV Hondius** expedition cruise ship. Cases first appeared in early April 2026; by the time infection was confirmed, passengers and crew had dispersed across Europe, Asia, and the Americas. Whole-genome sequencing confirms ANDV as the causative agent.

## Methods

Transmission trees were inferred with **outbreaker2**, integrating symptom onset dates and genome sequences. Serial interval was modelled as Gamma(mean = 18 d, CV = 1/3). MCMC: 100,000 iterations, thinned every 50 steps, 10% burn-in. L-segment sequences were retrieved from [Pathoplexus](https://pathoplexus.org) for the `r length(dna)` of `r nrow(linelist)` cases with deposited accessions. Remaining cases contribute temporal data only. Because near-zero genetic variation renders the per-site mutation rate ($\mu$) non-identifiable from the data, $\mu$ was fixed at $1.7 \times 10^{-5}$ substitutions/site/generation, derived from the published ANDV substitution rate of ${\sim}3.5 \times 10^{-4}$ subs/site/year and an 18-day generation interval.


## Results

### Timeline of Exposure Windows


```{r ship-timeline}
#| fig-cap: "Ship exposure windows per case. Bars coloured by role; circle = symptom onset; circle-X = death."
#| fig-height: 4.5
ctd_plot <- ctd_timed |>
  left_join(
    linelist |> select(gh_id, group, best_date, date_death, age, gender),
    by = c("id" = "gh_id")
  ) |>
  mutate(
    start_date = ref_date + start,
    end_date = ref_date + end,
    sex = str_to_upper(str_sub(gender, 1, 1)),
    label = fct_reorder(
      str_glue("{id}  ({sex}, {age})"),
      coalesce(end_date, start_date)
    )
  )

events <- ctd_plot |>
  select(label, Onset = best_date, Death = date_death) |>
  pivot_longer(c(Onset, Death), names_to = "event", values_to = "date") |>
  drop_na(date)

ggplot(ctd_plot) +
  geom_segment(
    aes(
      y = label,
      yend = label,
      x = start_date,
      xend = end_date,
      colour = group
    ),
    linewidth = 4,
    alpha = 0.75,
    lineend = "round"
  ) +
  geom_point(
    data = events,
    aes(y = label, x = date, colour = event, shape = event),
    size = 3,
    fill = "white",
    stroke = 1.4
  ) +
  scale_colour_manual(
    values = c(group_pal, Onset = "grey20", Death = "#c0392b"),
    labels = c(group_labels, Onset = "Symptom onset", Death = "Death"),
    breaks = c("passenger", "crew", "Onset", "Death"),
    name = NULL
  ) +
  scale_shape_manual(
    values = c(passenger = NA, crew = NA, Onset = 21, Death = 13),
    guide = "none"
  ) +
  scale_x_date(
    date_breaks = "1 week",
    date_labels = "%d %b",
    expand = expansion(add = 3)
  ) +
  labs(x = NULL, y = NULL) +
  theme_bw(base_size = 13) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 30, hjust = 1)
  ) +
  guides(
    colour = guide_legend(
      override.aes = list(
        linewidth = c(3, 3, NA, NA),
        linetype = c(1, 1, 0, 0),
        shape = c(NA, NA, 21, 13),
        size = c(0, 0, 3, 3),
        fill = c(NA, NA, "white", "white")
      )
    )
  )
```

### Genomic Analysis

```{r snp-matrix}
#| fig-cap: "Pairwise SNP distance matrix for sequenced cases (L segment). Values are raw nucleotide differences. Near-zero distances confirm a near-clonal cluster consistent with a single spillover event."
#| fig-height: 3.5
snp_dist <- ape::dist.dna(dna, model = "N", pairwise.deletion = TRUE) |>
  as.matrix()

pheatmap(
  snp_dist,
  color = colorRampPalette(c("#f7fbff", "#21b528"))(50),
  display_numbers = TRUE,
  number_format = "%.0f",
  number_color = "black",
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  treeheight_row = 20,
  treeheight_col = 20,
  border_color = "white",
  fontsize = 11,
  legend = FALSE,
  main = ""
)
```

### Model Diagnostics

```{r diagnostics-trace}
#| fig-cap: "MCMC trace plots for scalar parameters (post-burnin). Stationary, well-mixed traces indicate convergence."
#| fig-height: 7
param_labels <- c(
  post = "Log posterior",
  mu = "Mutation rate (μ)",
  pi = "Reporting rate (π)",
  eps = "Contact reporting coverage (ε)"
)

o2 |>
  mutate(chain = paste("Chain", chain)) |>
  select(chain, step, all_of(names(param_labels))) |>
  pivot_longer(-c(chain, step), names_to = "param", values_to = "value") |>
  mutate(param = factor(param_labels[param], levels = param_labels)) |>
  ggplot(aes(x = step, y = value, colour = chain)) +
  geom_line(linewidth = 0.25, alpha = 0.8) +
  facet_wrap(~param, scales = "free_y", ncol = 1) +
  scale_colour_brewer(palette = "Set1", name = NULL) +
  labs(x = "MCMC iteration", y = NULL) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "#f0f0f0"),
    legend.position = "bottom"
  )
```

```{r diagnostics-gelman}
#| tbl-cap: "Convergence diagnostics across 4 independent chains. R-hat < 1.01"
# Sampled parameters of the outbreaker2 model:
#   - pi (scalar, move_pi = TRUE)
#   - t_inf_i (per-case infection times, continuous; move_t_inf defaults to TRUE)
# Fixed parameters (mu, eps, tau) are omitted.
diag_cols <- c("pi", stringr::str_subset(names(o2), "^t_inf_\\d+$"))

mcmc_chains <- o2 |>
  group_split(chain) |>
  map(\(df) coda::mcmc(as.matrix(df[diag_cols]))) |>
  coda::mcmc.list()

gd  <- coda::gelman.diag(mcmc_chains, autoburnin = FALSE, multivariate = FALSE)
ess <- coda::effectiveSize(mcmc_chains)

n_chains  <- n_distinct(o2$chain)
n_samples <- nrow(o2)

tibble(
  param = names(ess),
  group = if_else(param == "pi", "pi", "t_inf"),
  rhat  = gd$psrf[, "Point est."],
  ess   = as.numeric(ess)
) |>
  group_by(group) |>
  summarise(
    n_params = n(),
    `R-hat`  = max(rhat),
    ESS      = round(min(ess)),
    .groups  = "drop"
  ) |>
  arrange(match(group, c("pi", "t_inf"))) |>
  mutate(
    Parameter = if_else(
      group == "pi",
      "Reporting rate (π)",
      str_c("Infection times (t_inf, N = ", n_params, ")")
    )
  ) |>
  select(Parameter, `R-hat`, ESS) |>
  gt() |>
  fmt_number(columns = `R-hat`, decimals = 3) |>
  fmt_integer(columns = ESS) |>
  tab_style(cell_fill("#d4edda"), cells_body(rows = `R-hat` < 1.01)) |>
  tab_style(cell_fill("#fff3cd"), cells_body(rows = `R-hat` >= 1.01 & `R-hat` < 1.05)) |>
  tab_style(cell_fill("#f8d7da"), cells_body(rows = `R-hat` >= 1.05)) |>
  tab_style(
    list(cell_fill("#2c3e50"), cell_text(color = "white", weight = "bold")),
    cells_column_labels()
  ) |>
  tab_footnote(str_glue(
    "ESS: effective sample size out of {n_samples} total posterior samples ",
    "({n_chains} chains × {n_samples / n_chains} samples per chain). ",
    "The t_inf row shows the worst-case (max R-hat, min ESS) across all per-case infection times."
  )) |>
  tab_options(table.width = pct(65), table.font.size = 13)
```

### Transmission Tree

```{r fig-network-static}
#| fig-cap: "Consensus transmission tree (coloured edges), where each case is assigned its most frequently inferred ancestor across the posterior sample. Grey dashed arrows show the posterior support for direct transmission from case 1 to all cases not already linked to it in the consensus tree, consistent with a superspreading event."
#| fig-height: 6
draw_key_arrow <- function(data, params, size) {
  grid::segmentsGrob(
    x0 = 0.1,
    x1 = 0.85,
    y0 = 0.5,
    y1 = 0.5,
    arrow = grid::arrow(length = unit(2, "mm"), type = "closed"),
    gp = grid::gpar(
      col = "grey25",
      fill = "grey25",
      lwd = (data$edge_width %||% 0.5) * ggplot2::.pt
    )
  )
}

epi <- make_epicontacts(
  linelist = linelist,
  contacts = consensus_tree |> drop_na(from),
  directed = TRUE
)

g <- epicontacts:::as.igraph.epicontacts(epi) |> as_tbl_graph()

# Reingold-Tilford: designed for rooted trees, guarantees no overlap
roots <- which(igraph::degree(g, mode = "in") == 0)
tree_layout <- igraph::layout_as_tree(g, root = roots)
rownames(tree_layout) <- igraph::V(g)$name

layout_data <- create_layout(g, layout = "kk")
layout_data$x <- as.numeric(layout_data$best_date)
layout_data$y <- tree_layout[layout_data$name, 1] # tree breadth → y axis

# Supporting edges: case "1" → every node not already linked in the consensus tree
already_linked <- consensus_tree |> filter(from == "1") |> pull(to)
ss_targets <- setdiff(linelist$gh_id, c("1", already_linked))
src <- layout_data |> filter(name == "1")

ss_segs <- tibble(
  x = src$x,
  y = src$y,
  xend = layout_data$x[match(ss_targets, layout_data$name)],
  yend = layout_data$y[match(ss_targets, layout_data$name)],
  freq = map_dbl(ss_targets, \(t) {
    mean(o2_id[[str_c("alpha_", t)]] == "1", na.rm = TRUE)
  })
) |>
  mutate(xmid = (x + xend) / 2, ymid = (y + yend) / 2)

ggraph(layout_data) +
  geom_segment(
    data = ss_segs,
    aes(x = x, y = y, xend = xend, yend = yend),
    colour = "grey60",
    linetype = "dashed",
    arrow = arrow(length = unit(2.5, "mm"), type = "closed"),
    inherit.aes = FALSE
  ) +
  geom_text(
    data = ss_segs,
    aes(x = xmid, y = ymid, label = sprintf("%.2f", freq)),
    colour = "grey50",
    size = 3,
    inherit.aes = FALSE
  ) +
  geom_edge_link(
    aes(
      edge_width = frequency,
      color = .N()$group[from],
      label = sprintf("%.2f", frequency)
    ),
    arrow = arrow(length = unit(2.5, "mm")),
    end_cap = circle(3, "mm"),
    angle_calc = "along",
    label_dodge = unit(2.5, "mm"),
    label_size = 3,
    label_colour = "black",
    key_glyph = draw_key_arrow
  ) +
  geom_node_point(aes(fill = group), shape = 21, colour = "black", size = 5) +
  geom_node_text(
    aes(label = name),
    repel = TRUE,
    size = 3,
    nudge_y = 0.1,
    bg.color = "white",
    bg.r = 0.1
  ) +
  scale_edge_width(
    "Posterior support",
    range = c(0.1, 1),
    breaks = c(0.8, 0.9, 1)
  ) +
  scale_fill_manual(
    NULL,
    values = group_pal,
    breaks = names(group_pal),
    labels = group_labels
  ) +
  scale_edge_colour_manual(values = group_pal, guide = "none") +
  guides(
    fill = guide_legend(override.aes = list(size = 4), order = 1),
    edge_width = guide_legend(order = 2)
  ) +
  scale_x_continuous(breaks = breaks_width(3), labels = \(x) {
    format(as.Date(x, origin = "1970-01-01"), "%b\n%d")
  }) +
  labs(x = "Date (onset / confirmation / hospitalisation)", y = "") +
  theme_bw(base_size = 13) +
  theme(
    axis.line.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "bottom"
  )
```

### Ancestry Uncertainty

```{r entropy-plot}
#| fig-cap: "Shannon entropy of each case's inferred source. Entropy = 0 means a single unambiguous infector."
#| fig-height: 4
onset_order <- linelist |> arrange(desc(dates)) |> pull(gh_id)

tibble(
  case = gsub("alpha_", "", names(entropy_vals)),
  entropy = entropy_vals
) |>
  mutate(case = factor(case, levels = onset_order)) |>
  ggplot(aes(x = entropy, y = case, fill = entropy)) +
  geom_col() +
  scale_fill_gradient(low = "#caced1", high = "#383939", guide = "none") +
  labs(x = "Shannon entropy", y = "Case ID") +
  theme_bw(base_size = 13) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )
```

### Dates of Infection

```{r infection-dates}
#| fig-cap: "Posterior distribution of inferred infection dates per case. Width reflects uncertainty; cases ordered by mean inferred infection date."
#| fig-height: 6
o2 |>
  select(starts_with("t_inf_")) |>
  pivot_longer(
    everything(),
    names_to = "idx",
    names_pattern = "t_inf_(\\d+)",
    names_transform = list(idx = as.integer),
    values_to = "t_inf"
  ) |>
  mutate(
    case = fct_reorder(linelist$gh_id[idx], t_inf, mean, .desc = TRUE),
    grp = linelist$group[idx],
    date = ref_date + t_inf
  ) |>
  ggplot(aes(x = date, y = case, fill = grp)) +
  geom_density_ridges(alpha = 0.75, colour = "white", linewidth = 0.4) +
  scale_fill_manual(
    values = group_pal,
    na.value = "grey70",
    labels = group_labels,
    name = NULL
  ) +
  scale_x_date(
    date_breaks = "7 days",
    date_labels = "%d %b",
    expand = expansion(add = 3)
  ) +
  labs(x = "Inferred date of infection", y = NULL) +
  theme_bw(base_size = 13) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 30, hjust = 1)
  )
```

### Imports and Unobserved Transmission

```{r import-kappa}
#| fig-cap: "Posterior probability of each case being imported, directly linked, or connected via unobserved intermediate cases (κ)."
#| fig-height: 5
state_pal <- c(
  "Imported" = "#e6550d",
  "Direct (κ = 1)" = "#525252",
  "1 unobserved (κ = 2)" = "#817f7f",
  "2+ unobserved (κ ≥ 3)" = "#c4c3c3"
)

case_order <- linelist |> arrange(desc(best_date)) |> pull(gh_id)

o2 |>
  select(matches("^(alpha|kappa)_\\d+$")) |>
  pivot_longer(
    everything(),
    names_to = c(".value", "idx"),
    names_pattern = "(alpha|kappa)_(\\d+)",
    names_transform = list(idx = as.integer)
  ) |>
  mutate(
    case = factor(linelist$gh_id[idx], levels = case_order),
    state = factor(
      case_when(
        is.na(alpha) ~ "Imported",
        kappa == 1 ~ "Direct (κ = 1)",
        kappa == 2 ~ "1 unobserved (κ = 2)",
        .default = "2+ unobserved (κ ≥ 3)"
      ),
      levels = names(state_pal)
    )
  ) |>
  count(case, state) |>
  group_by(case) |>
  mutate(prob = n / sum(n)) |>
  ungroup() |>
  complete(case, state, fill = list(n = 0, prob = 0)) |>
  ggplot(aes(x = prob, y = case, fill = state)) +
  geom_col() +
  scale_x_continuous(
    labels = percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.02))
  ) +
  scale_fill_manual(values = state_pal, name = NULL) +
  labs(x = "Posterior frequency", y = NULL) +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

### Reproduction Number

```{r offspring}
#| fig-cap: "Posterior distribution of the offspring distribution."
#| fig-height: 4
get_Ri(o2_id) |>
  mutate(step = row_number()) |>
  pivot_longer(-step, names_to = "name", values_to = "Ri") |>
  count(step, Ri) |>
  group_by(step) |>
  mutate(percent = n / sum(n) * 100) |>
  group_by(Ri) |>
  summarise(
    mean = mean(percent),
    lower = quantile(percent, 0.025),
    upper = quantile(percent, 0.975),
    .groups = "drop"
  ) |>
  ggplot(aes(x = Ri, y = mean)) +
  geom_linerange(
    aes(ymin = 0, ymax = mean),
    colour = "grey60",
    linewidth = 0.5
  ) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
  theme_bw()
```


## Note
Cases 23 (WHO Case 12, Dutch crew member) and 24 (WHO Case 13, Spanish national) are inferred as imported by the model. Both lack symptom onset dates — the model falls back to their confirmation dates (22 and 25 May), placing them 49 and 52 days after the index case. At those dates, the serial interval likelihood of linkage to any earlier case is very low. Neither case has a sequenced sample, removing the genetic signal that would otherwise support a specific ancestry. Their ship disembarkation dates are also absent from the linelist, so the shared-location contact window contributes no additional evidence. Combined with the aggressive import threshold (`outlier_threshold = 2.5`), the model consistently assigns `α = NA` to both cases. They may well be secondary cases within the ship chain; the classification would likely change if onset dates or sequences become available.

## Data

Data available under [CC BY 4.0](https://creativecommons.org/licenses/by/4.0/) at [kraemer-lab/Hondius_hantavirus_h2026](https://github.com/kraemer-lab/Hondius_hantavirus_h2026). Sequences retrieved from [Pathoplexus](https://pathoplexus.org). Report source: `report.qmd`.
